### 2020-02-01 background.tex
some more misc changes, still to do cv and nlp sections

Statistical mechanical ideas have long had influence on NN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (i.e., those using fitted PL exponents and HT-SR Theory) are based on statistical mechanics ideas~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}.
However, that the way in which we \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We will \emph{use} statistical mechanics theory more as it is used in quantitative finance.
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statisical mechanics of heavy tailed and strongly correlated systems.

>>>>

Statistical mechanical ideas have long had influence on NN theory and practice~\cite{EB01_BOOK, MM17_TR, BKPx20}; and 
our best-performing metrics (i.e., those using fitted PL exponents and HT-SR Theory) are based on statistical mechanics ideas~\cite{MM17_TR, MM18_TR, MM19_HTSR_ICML, MM19_KDD, MM20_SDM}.
\michael{Where first define HT-SR.}
However, that the way in which we \emph{use} statistical mechanics theory is quite different than the way it is more commonly formulated.
Several very good overviews of the more common approach are available~\cite{EB01_BOOK, BKPx20}.
We will \emph{use} statistical mechanics theory more as it is used in quantitative finance.
Thus, much more relevant for our metholodogical approach is older work of Bouchaud, Potters, Sornette, and coworkers~\cite{BouchaudPotters03, SornetteBook, BP11, bun2017} on the statisical mechanics of heavy tailed and strongly correlated systems.

### 2020-02-01 cv_models.tex
some more misc changes, still to do cv and nlp sections

In this section, we examine quality metrics described in Section~\ref{sxn:methods} for several Computer Vision (CV) model architecture series.
This includes the VGG, ResNet, and DenseNet series of models, each of which consists of several pretrained DNN models, trained on the full ImageNet~\cite{imagenet} dataset, and each of which is distributed with the current opensource pyTorch framework (version 1.4)~\cite{pyTorch}.
This also includes a larger set of ResNet models, trained on the ImageNet-1K dataset~\cite{imagenet1k}, provided on the OSMR ``Sandbox for training convolutional networks for computer vision''~\cite{osmr}, which we call the ResNet-1K series.

We use our quality metrics to perform \emph{coarse model analysis}, comparing and contrasting the three model series (or four, if ResNet-1K is included), permitting us to predict trends in model quality. 
We also use perform a more \emph{fine layer analysis} as a function of depth for these models, providing novel insights among the VGG, ResNet, and DenseNet architectures.

@@

We have examined the performance of the four quality metrics (Frobenius norm, Spectral norm, Weighted Alpha, and $\alpha$-Norm) applied to each of the VGG, ResNet, ResNet-1K, and DenseNet series.
To start, Figure~\ref{fig:vgg-metrics} considers the VGG series (in particular, the pretrained models VGG11, VGG13, VGG16, and VGG19, with and without BatchNormalization), and it plots the four quality metrics versus the reported Test accuracies~\cite{pyTorchVgg}, as well as a basic linear regression line. 
%with the Root Mean Squared Error (RMSE) shown, \michael{XXX.  WHERE IS RMSE.  IN THAT TABLE?}
All four metrics correlate quite well with the reported Top1 Accuracies, with smaller norms and smaller values of $\hat{\alpha}$ implying better generalziation (i.e., greater accuracy, lower error). 
While all four metrics perform well, notice that the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) performs best (with an RMSE of $0.42$, see Table~\ref{table:cv-models}); and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $), which is an approximation to the Log $\alpha$-Norm metric~\cite{MM20_unpub_work}, performs second best (with an RMSE of $0.48$, see Table~\ref{table:cv-models}).

>>>>

In this section, we examine quality metrics described in Section~\ref{sxn:methods} for several CV model architecture series.
This includes the VGG, ResNet, and DenseNet series of models, each of which consists of several pretrained DNN models, trained on the full ImageNet~\cite{imagenet} dataset, and each of which is distributed with the current opensource pyTorch framework (version 1.4)~\cite{pyTorch}.
This also includes a larger set of ResNet models, trained on the ImageNet-1K dataset~\cite{imagenet1k}, provided on the OSMR ``Sandbox for training convolutional networks for computer vision''~\cite{osmr}, which we call the ResNet-1K series.

We perform \emph{coarse model analysis}, comparing and contrasting the four model series, and predicting trends in model quality. 
We also perform \emph{fine layer analysis}, as a function of depth for these models, illustrating that PL-based metrics can provide novel insights among the VGG, ResNet/ResNet-1K, and DenseNet architectures.

@@

We have examined the performance of the four quality metrics (Log Frobenius norm, Log Spectral norm, Weighted Alpha, and Log $\alpha$-Norm) applied to each of the VGG, ResNet, ResNet-1K, and DenseNet series.
To start, Figure~\ref{fig:vgg-metrics} considers the VGG series (in particular, the pretrained models VGG11, VGG13, VGG16, and VGG19, with and without BatchNormalization), and it plots the four quality metrics versus the reported Test accuracies~\cite{pyTorchVgg}, as well as a basic linear regression line. 
%with the Root Mean Squared Error (RMSE) shown, \michael{XXX.  WHERE IS RMSE.  IN THAT TABLE?}
All four metrics correlate quite well with the reported Top1 Accuracies, with smaller norms and smaller values of $\hat{\alpha}$ implying better generalziation (i.e., greater accuracy, lower error). 
While all four metrics perform well, notice that the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) performs best (with an RMSE of $0.42$, see Table~\ref{table:cv-models}); and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $), which is an approximation to the Log $\alpha$-Norm metric~\cite{MM20_unpub_work}, performs second best (with an RMSE of $0.48$, see Table~\ref{table:cv-models}).

### 2020-02-01 methods.tex
some more misc changes, still to do cv and nlp sections

Each DNN layer contains one or more layer 2D weight matrices $\mathbf{W}_{l}$, and/or 2D feature maps $\mathbf{W}_{i,l}$ extracted from 2D Convolutional layers. 
(For notational convenience, we may drop the $i$ and/or $i,l$ subscripts below; and we let $\mathbf{W}$ be a generic $N\times M$ weight matrix, with $N\ge M$.) 
%
We have examined a large number of possible quality metrics.
The best performing metrics (recall that we can only consider metrics that do not use training/test data) depend on the norm and/or spectral properties of weight matrices, $\mathbf{W}$.%
\footnote{We do not use intra-layer information from the models in our quality metrics, but as we will describe our metrics can be used to learn about that.}
We consider the following metrics.
\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\sum_{i,j}w^{2}_{i,j} = \sum_{i=1}^{M} \lambda_{i}^{2}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}=\lambda_{max}$
\michael{Should this be $\sqrt{\lambda_{max}}$?}
\item 
Weighted Alpha Metric: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
The first two metrics are well-known in ML.
The last two deserve special mention.
For all these metrics, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the \emph{Empirical Correlation Matrix},
$ %% $$
\mathbf{X}=\mathbf{W}^{T}\mathbf{W} ,
$ %% $$
and so $\lambda_{max}$ is the maximum eigenvalue of $\mathbf{X}$. 
(These eigenvalues are the squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, i.e., $\lambda_{i}=\sigma^{2}_{i}$.)
For the last two metrics, $\alpha$ is the PL exponent that arises in the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Operationally, $\alpha$ is determined by using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}  ,
\end{equation}
where $\lambda_{max}$ \emph{is} the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
The Weighted Alpha Metric was introduced previously~\cite{MM20_SDM}, where (on a much smaller set of data than we consider here) it was shown to correlate well with the trends in reported test accuracies of pretrained DNNs.
Based on this, here we introduce and evaluate the $\alpha$-Norm metric.
One would expect $\hat{\alpha}$ to approximate the log $\alpha$-Norm very well for $\alpha < 2$ and reasoably well for $\alpha\in[2,5]$~\cite{MM20_unpub_work}.

Each of these metrics is defined for a given layer $\mathbf{W}$ matrix.
For the Weighted Alpha Metric, we consider the average of $\hat{\alpha}$ over all layers, i.e., $\sum_l \hat{\alpha}_l = \sum_l \alpha_l\log\lambda_{max,l} $ .  
\michael{Be careful to get that right.}
For the norm-based metrics, we consider the \emph{weighted} average of the log of the norm to the appropriate power.
For example, for the $\alpha$-Norm, we compute 
\begin{equation}
\label{eqn:sum_log_alpha_norm_alpha}
\sum\nolimits_l \log \Vert\mathbf{W}_l\Vert_{\alpha_l}^{\alpha_l} 
=
\sum\nolimits_l \alpha_l \log \Vert\mathbf{W}_l\Vert_{\alpha_l} .
\end{equation}
Informallly, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
In this case, the log Complexity takes the form of an Average Log Norm,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert  .
\end{equation}
For the Frobenius and Spectral norms, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly. 
For example, when taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor.
For the Weighted Alpha Norm, however, $\alpha_l$ varies from layer to layer, and so in Eqn.~(\ref{eqn:sum_log_alpha_norm_alpha}) it cannot be taken out of the sum.

>>>>

Each DNN layer contains one or more layer 2D weight matrices $\mathbf{W}_{l}$, and/or 2D feature maps $\mathbf{W}_{i,l}$ extracted from 2D Convolutional layers. 
(For notational convenience, we may drop the $i$ and/or $i,l$ subscripts below; and we let $\mathbf{W}$ be a generic $N\times M$ weight matrix, with $N\ge M$.) 
%
We have examined a large number of possible quality metrics.
The best performing metrics (recall that we can only consider metrics that do not use training/test data) depend on the norm and/or spectral properties of weight matrices, $\mathbf{W}$.%
\footnote{We do not use intra-layer information from the models in our quality metrics, but as we will describe our metrics can be used to learn about that.}
Consider the following.
\begin{itemize}
\item 
Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\sum_{i,j}w^{2}_{i,j} = \sum_{i=1}^{M} \lambda_{i}^{2}$
\item 
Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}=\lambda_{max}$
\michael{Should this be $\sqrt{\lambda_{max}}$?}
\item 
Weighted Alpha Metric: $\hat{\alpha}=\alpha\log\lambda_{max}$
\item 
$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{\alpha}_{\alpha}=\sum_{i=1}^{M}\lambda_{i}^{\alpha}$
\end{itemize}
The first two quantities are well-known in ML.
The last two deserve special mention.
For all these quantities, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the \emph{Empirical Correlation Matrix},
$ %% $$
\mathbf{X}=\mathbf{W}^{T}\mathbf{W} ,
$ %% $$
and so $\lambda_{max}$ is the maximum eigenvalue of $\mathbf{X}$. 
(These eigenvalues are the squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, i.e., $\lambda_{i}=\sigma^{2}_{i}$.)
For the last two quantities, $\alpha$ is the PL exponent that arises in the recently-developed \emph{Theory of Heavy Tailed Self Regularization (HT-SR)}~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
Operationally, $\alpha$ is determined by using the publicly-available \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the Empirical Spectral Density (ESD) of $\mathbf{X}$, i.e., a histogram of the eigenvalues, call it $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\le\lambda_{max}  ,
\end{equation}
where $\lambda_{max}$ \emph{is} the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.

For norm-based metrics, we consider the average log norm to the appropriate power.
For example, for the \emph{$\alpha$-Norm metric}, 
\begin{equation}
\label{eqn:sum_log_alpha_norm_alpha}
\sum\nolimits_l \log \Vert\mathbf{W}_l\Vert_{\alpha_l}^{\alpha_l} 
=
\sum\nolimits_l \alpha_l \log \Vert\mathbf{W}_l\Vert_{\alpha_l} .
\end{equation}
Informallly, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
The log Complexity,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert  ,
\end{equation}
 takes the form of an Average Log Norm.
For the \emph{Frobenius norm metric} and \emph{Spectral norm metric}, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly. 
(When taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor.)
For the Weighted Alpha Norm, however, $\alpha_l$ varies from layer to layer, and so in Eqn.~(\ref{eqn:sum_log_alpha_norm_alpha}) it cannot be taken out of the sum.

For the Weighted Alpha Metric, we consider the average of $\hat{\alpha}$ over all layers, i.e., 
$ %$$
\sum_l \hat{\alpha}_l = \sum_l \alpha_l\log\lambda_{max,l}  .  
$ %$$
\michael{Be careful to get that right.}
The Weighted Alpha Metric was introduced previously~\cite{MM20_SDM}, where (on a much more limited set of data than we consider here) it was shown to correlate well with the trends in reported test accuracies of pretrained DNNs.
Based on this, in this paper, we introduce and evaluate the $\alpha$-Norm metric.
One would expect $\hat{\alpha}$ to approximate the log $\alpha$-Norm very well for $\alpha < 2$ and reasoably well for $\alpha\in[2,5]$~\cite{MM20_unpub_work}.
